Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
fix a few of the fixes/remarks

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning \anto{this is not true: what if we trace the last iteration?}. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

@@

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/get} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

Note how the functions that are called by \texttt{f} are automatically inlined
into the trace. The method calls are always preceded by a \texttt{guard\_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
class check, not checking for subclasses} These guards make the trace specific
to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
already be said to be specialized for \texttt{BoxedIntegers}. When the trace is
turned into machine code and then executed with \texttt{BoxedFloats}, the
first \texttt{guard\_class} instruction will fail and execution will continue
using the interpreter.

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the result of 
the guard is statically known.

@@

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
whose \texttt{intval} field contains the constant -100.
\anto{this works only because we use SSI and thus the value of $i_{4}$ never
changes. However, SSI is not explained anywhere in the paper}

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{get} operations
and \texttt{guard\_class} operations).  Moreover, in real life example usually
the loops are more complex and contain more objects of type 1, thus this
technique is more effective.

@@

For the former we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The results can be
seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
little as XXX percent of allocation operations in the benchmarks. All benchmarks
taken together, the optimization removes XXX percent of allocation operations.
\anto{Actually, we can only know how many operations we removed from the traces, not from the actual execution}

>>>>

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

@@

\begin{figure}
\texttt{
\begin{tabular}{l}
\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res \\
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

If the function is executed using the tracing JIT, with \texttt{y} being a
\texttt{BoxedInteger}, the produced trace looks like
Figure~\ref{fig:unopt-trace}. The operations in the trace are shown indented to
correspond to the stack level of the function that contains the traced
operation. The trace is in single-assignment form, meaning that each variable is
assigned to exactly once. The trace also shows the inefficiencies of \texttt{f} clearly, if one
looks at the number of \texttt{new} (corresponding to object creation),
\texttt{set/get} (corresponding to attribute reads/writes) and
\texttt{guard\_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

Note how the functions that are called by \texttt{f} are automatically inlined
into the trace. The method calls are always preceded by a \texttt{guard\_class}
operation, to check that the class of the receiver is the same as the one that
was observed during tracing.\footnote{\texttt{guard\_class} performs a precise
class check, not checking for subclasses.} These guards make the trace specific
to the situation where \texttt{y} is really a \texttt{BoxedInteger}, it can
already be said to be specialized for \texttt{BoxedIntegers}. When the trace is
turned into machine code and then executed with \texttt{BoxedFloats}, the
first \texttt{guard\_class} instruction will fail and execution will continue
using the interpreter.

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check che guard does is statically known.

@@

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
whose \texttt{intval} field contains the constant -100.

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{get} operations
and \texttt{guard\_class} operations).

@@

For the former we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The results can be
seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
little as XXX percent of allocation operations in the traces of the benchmarks.
All benchmarks taken together, the optimization removes XXX percent of
allocation operations.
